---
title: "DA5020 - Week 6 Assignment Tidy and Relational Data Operations"
output:
  pdf_document: default
  word_document: default
date: '`r Sys.Date()`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  # mute messages output
  message = FALSE
)
```

Packages needed to complete analysis:
tidyverse (dplyr)

This week's assignment is about tidying up the structure of data collected by the US census. Load the Unemployment and Educational data files into R studio. One file contains yearly unemployment rates from 1970 to 2015, for counties in the US. The other file contains aggregated data percentages on the highest level of education achieved for each census member. The levels of education are: "less than a high school diploma", "high school diploma awarded", "attended some college", "college graduate and beyond". The census tracks the information at the county level and uses a fips number to represent a specific county within a U.S. state. The fips number is a 5 digit number where the first two digits of the fips number represents a U.S. state, while the last three digits represent a specific county within that state.  


## Questions

1. (20 points) Download the unemployment and education data files from blackboard and save the files to your working directory folder. Load both the unemployment data and the education data into R. Review the education data. Identify where variable names are actually values for a specific variable. Identify when multiple rows are data for the same entity. Identify when specific columns contain more than one atomic value. Tidy up the education data using spread, gather and separate. 

# Load in separate files for unemployment data (renamed to unemploymentData) and 
# education data (renamed to educationData)
```{r, eval = FALSE, echo = TRUE}
unemploymentData <- read_csv("FipsUnemploymentDA5020.csv")

educationData <- read_csv("FipsEducationsDA5020.csv")
```

# Tidy up educationData, first by spreading the 'percent_measure' column to give each
# meaningful variable its own column. Then separate the 'county_state' column to give proper
# representation of the two
```{r, eval = FALSE, echo = TRUE}
educationData$county_state <- str_replace_all(educationData$county_state, " (\\|) ", "_")
educationData <- educationData %>%
  spread(key = percent_measure, value = percent) %>%
  separate(county_state, c("state", "county"), "_")
```


2. (15 points) Break apart the education data into three distinct tibbles. One tibble named education contains the education data, another tibble named fips, contains the fips number definition, and the third tibble named rural_urban_code contains the textual description of the 9 different urban to rural data descriptions. These three tibbles must be linked together to represent the relationships between the tibbles. For example, the fips table will contain 3,192 rows, where each row represents the definition of a fips number (County, State). Each row in the education table will contain the educational attainment of a spcific county. It also will contain a fips number since this data is specific to a county within a state. 


```{r, eval = FALSE, echo = TRUE}
education <- tibble(
  `county` = educationData$county,
  `fips` = educationData$fips,
  `year` = educationData$year,
  `percent_four_plus_years_college` = educationData$percent_four_plus_years_college,
  `percent_has_some_college` = educationData$percent_has_some_college,
  `percent_hs_diploma` = educationData$percent_hs_diploma,
  `percent_less than_hs_diploma` = educationData$`percent_less than_hs_diploma`
)

 fips <- tibble(
 `fips` = educationData$fips,
 `state` = educationData$state,
 `county` = educationData$county
) %>%
   unique()

rural_urban_code <- tibble(
  `rural_urban_cont_code` = educationData$rural_urban_cont_code,
  `description` = educationData$description
) %>%
  unique() %>%
  arrange(`rural_urban_cont_code`)
```


3. (5 points) Answer the following questions about your tibbles: The fips column in the education table - is it a foreign or a primary key for the education tibble?  What is the primary key for your education tibble? The rural_urban code tibble should only contain 9 rows. What is its primary key?

# The 'fips' column is a foreign key for the education tibble.
# There is no clear primary key for my education tibble. A surrogate should be created.
# The primary key for the rural_urban_code tibble is the 'Code' variable. Each value in the 
# key uniquely identifies a observation within the dataset.


4. (50 points) Write expressions to answer the following queries: 

4.0 In the year 1970, what is the percent of the population not attaining a high school diploma for the Nantucket county in Massachusetts? What about the year 2015?

```{r, eval = FALSE, echo = TRUE}
education %>%
  select(county, year, `percent_less than_hs_diploma`) %>%
  filter(county == "Nantucket County", year == 1970)

education %>%
  select(county, year, `percent_less than_hs_diploma`) %>%
  filter(county == "Nantucket County", year == 2015)
```

4.1 What is the average percentage not receiving a high school diploma for the counties in Alabama for the year 2015?

```{r, eval = FALSE, echo = TRUE}
education %>%
  select(fips, county, year, `percent_less than_hs_diploma`) %>%
  filter(between(fips, 1001, 1133), year == 2015) %>%
  summarise(
    "Average Percentage Not Receiving HS Diploma for Counties of Alabama in 2015" = mean(
      `percent_less than_hs_diploma`, na.rm = TRUE)
  )
```

4.2 What is the average percentage of college graduates for the counties in the state of Massachusetts for the year 2015?

```{r, eval = FALSE, echo = TRUE}
education %>%
  select(fips, county, year, percent_four_plus_years_college) %>%
  filter(between(fips, 25001, 25027), year == 2015) %>%
  summarise(
    "Average Percentage of College Graduates for Counties of Massachusetts in 2015" = mean(
      percent_four_plus_years_college, na.rm = TRUE)
  )
```

4.3 Determine the average percentage of population not attaining a high school diploma for the counties in Alabama for each year within the dataset. The result should return the calendar year and the average percentage not attaining a high school diploma for that year.

```{r, eval = FALSE, echo = TRUE}
education %>%
  select(fips, county, year, `percent_less than_hs_diploma`) %>%
  filter(between(fips, 1001, 1133)) %>%
  group_by(year) %>%
  summarise(
    "Average Percentage No Receiving HS Diploma for Counties of Alabama" = mean(
      `percent_less than_hs_diploma`
    )
  )
```

4.4 What is the most common rural_urban code for the U.S. counties?

```{r, eval = FALSE, echo = TRUE}
educationData %>%
  count(rural_urban_cont_code) %>%
  arrange(desc(n))

rural_urban_code %>% filter(rural_urban_cont_code == 6)
```

4.5 Which counties have not been coded with a rural urban code? Return a result that contains two fields: County, State for the counties that have not been assigned a rural urban code. Do not return duplicate values in the result. Order the result alphabetically by state. What does this result set represent?

# These results represent the entries that have been coded to the whole state therefore
# there is no individual rural urban code.
```{r, eval = FALSE, echo = TRUE}
educationData %>%
  group_by(county, state) %>%
  filter(rural_urban_cont_code == "NULL") %>%
  select(county, state) %>%
  arrange() %>%
  unique()
```

4.6 What is the minimal percentage of college graduates for the counties in the state of Mississippi for the year 2010? What does the result represent?

# Since the code returns an empty tibble with 0 rows, this result represents that 
# Mississippi did not submit data for the year 2010. 
```{r, eval = FALSE, echo = TRUE}
educationData %>%
  select(fips, state, county, year, percent_four_plus_years_college) %>%
  filter(between(fips, 28001, 28163), year == 2010)
```

4.7 In the year 2015, which fip counties, are above the average unemployment rate? Provide the county name, U.S. state name and the unemployment rate in the result. Sort in descending order by unemployment rate. 

```{r, eval = FALSE, echo = TRUE}
unemploymentByCounty2015 <- left_join(unemploymentData, fips, by = "fips") %>%
  filter(year == 2015) %>%
  select(county, state, percent_unemployed)

calculatedAverageUnemploymentRate2015 <- mean(unemploymentByCounty2015$percent_unemployed)

unemploymentByCounty2015 %>%
  filter(percent_unemployed > calculatedAverageUnemploymentRate2015) %>%
  arrange(desc(percent_unemployed))
```

4.8 In the year 2015, which fip counties, U.S. states contain a higher percentage of unemployed citizens than the percentage of college graduates? List the county name and the state name. Order the result alphabetically by state.

```{r, eval = FALSE, echo = TRUE}
unemployedVsCollegeGrads <- 
  left_join(unemploymentByCounty2015, education, by = "county") %>%
  filter(year == 2015, (percent_unemployed > percent_four_plus_years_college)) %>%
  select(county, state)
```

4.9 Return the county, U.S. state and year that contains the highest percentage of college graduates in this dataset?

```{r, eval = FALSE, echo = TRUE}
maxPercentageCollegeGrads <- max(educationData$percent_four_plus_years_college)

educationData %>% group_by(county, state, year) %>%
  filter(percent_four_plus_years_college == maxPercentageCollegeGrads) %>%
  select(county, state, year)
```

5. (10 points) _Open question_: explore the unemployment rate and the percent not attaining a high school diploma over the time period in common for the two datasets. What can you discover? Create a plot that supports your discovery.

```{r, eval = FALSE, echo = TRUE}
unemploymentOverYears <- unemploymentData %>%
  select(year, percent_unemployed) %>%
  group_by(year) %>%
  summarise(
    "Average Percent Unemployment" = mean(percent_unemployed, na.rm = TRUE)
  )

percentNoHSDiplomaOverYears <- education %>%
  select(year, `percent_less than_hs_diploma`) %>%
  group_by(year) %>%
  summarise(
    "Average Percent No HS Diploma" = mean(`percent_less than_hs_diploma`)
  )

unemploymentVsNoHSDiploma <- 
  inner_join(percentNoHSDiplomaOverYears, unemploymentOverYears, by = "year")

# After matching the unemployment rate to the percent of no HS diploma, initial belief would
# be that the percent of no HS diploma contributes heavily to the unemployment rate but if
# you look at the fluxation of the unemployment rate vs the steady decrease in percent with
# no HS diploma, you can see that this is not the case

ggplot(unemploymentOverYears) +
  geom_point(aes(x = year, y = `Average Percent Unemployment`))

ggplot(percentNoHSDiplomaOverYears) +
  geom_point(aes(x = year, y = `Average Percent No HS Diploma`))
```
