---
title: "Homework 2"
author: "Patrick Blaney"
date: "January 21, 2018"
output: pdf_document
---

Packages required for completion of assignment:
tidyverse
gapminder

Load and assign the CSV file of data needed to complete assignment
```{r, echo=TRUE}
surveys <- read.csv("surveys.csv", header = T, sep = ",")
```

This loads the surveys.csv data into the surveys data frame. The fields in the data table are the following:
record_id : a unique number for each row in the table
month : Month when the observation was made
day: Calendar day the observation was made
year: Year the observation was made
plot_id: the area the measurement was taken
species_Id: species id, please see https://github.com/weecology/portal-teachingdb/blob/master/species.csv for more information
sex: sex of the observation Male or Female
hindfoot: length of the hindfoot
weight: weight of the animal

Question 1
The code below filters the survey to display on the first 3 months of the year 1990.
```{r}
jan_feb_mar_1990 <- filter(surveys, year == 1990, month %in% c(1, 2, 3))
```

Question 2
The code below sorts the record ID by decending order of the winter months (December, January, February) in the year 1990.
```{r}
winter_1990_record_id <- surveys %>%
  group_by(year = 1990) %>%
  filter(month %in% c(12, 1, 2)) %>%
  arrange(desc(record_id), .by_group = TRUE)
```

The code below sorts the weight by ascending order of the winter months (December, January, February) in the year 1990
```{r}
winter_1990_weight <- surveys %>%
  group_by(year = 1990) %>%
  filter(month %in% c(12, 1, 2)) %>%
  arrange(weight, .by_group = TRUE)
```

Question 3
The code below shows record id, sex, and weight of all Reithrodontomys montanus
```{r}
reithrodontomys_montanus <- surveys %>%
  filter(species_id == "RO") %>%
  select(record_id, sex, weight)
```

Question 4
The code below shows the average weight and hindfoot length of all Dipodomys merriami for each month
```{r}
dipodomys_merriami <- surveys %>%
  filter(species_id == "DM") %>%
  group_by(month) %>%
  summarise(
    avg_weight = mean(weight, na.rm = TRUE),
    avg_hindfoot_length = mean(hindfoot_length, na.rm = TRUE)
  )
```

Question 5
The code below gives the total count of different species in the winter (December, January, February) of the year 1990
```{r}
species_count_winter_1990 <- surveys %>%
  group_by(year = 1990) %>%
  filter(month %in% c(12, 1, 2)) %>%
  summarise(
    number_of_diff_species = n_distinct(species_id, na.rm = TRUE)
  )
```

Question 6
The code below creates a dataframe of the gapminder data and adds a column which outputs the gross domestic product (GDP) for each row.
```{r}
(gapminder_df <- as.data.frame(gapminder) %>%
  mutate(GDP = gdpPercap * pop))
```

Question 7
The code below calculates the mean GDP for Cambodia for the years within the dataset
```{r}
cambodia_mean_gdp <- gapminder_df %>%
  filter(country == "Cambodia") %>%
  summarise(
    Mean_GDP = mean(GDP, na.rm = TRUE)
  )
```

Question 8
The code below determines the maximum life expectancy for all countries in Asia then arranges them in decending order by year.
```{r}
life_expectancy_asia <- gapminder_df %>%
  filter(continent == "Asia") %>%
  group_by(country) %>%
  summarise(
    max_life_expectancy = max(lifeExp),
    year_of_max = max(year)
  )
```

Question 9
The code below counts the number of observed data points per continent
```{r}
observations_per_continent <- gapminder_df %>%
  group_by(continent) %>%
  summarise(
    observations = n_distinct(gdpPercap)
  )
```

Question 10
The code below computes the average and median life expectancy and GDP per capita for each continent for the years 1952 and 2007
```{r}
optimisim_trend_1952 <- gapminder_df %>%
  filter(year == 1952) %>%
  group_by(continent) %>%
  summarise(
    avg_life_expectancy = mean(lifeExp, na.rm = TRUE),
    median_life_expectancy = median(lifeExp, na.rm = TRUE),
    avg_gdp_per_capita = mean(gdpPercap, na.rm = TRUE),
    median_gdp_per_capita = median(gdpPercap, na.rm = TRUE)
  )
```

```{r}
optimisim_trend_2007 <- gapminder_df %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarise(
    avg_life_expectancy = mean(lifeExp, na.rm = TRUE),
    median_life_expectancy = median(lifeExp, na.rm = TRUE),
    avg_gdp_per_capita = mean(gdpPercap, na.rm = TRUE),
    median_gdp_per_capita = median(gdpPercap, na.rm = TRUE)
  )
```
 
Based on the increase across the average life expectancy and average GDP per capita, we should be optimistic. Although, there are factors that play into both statistics reaching a plateau as years go on and we may be reaching that point. 