---
title: "R Notebook"
output: html_notebook
---

Source Code for Homework 7B

# First steps consist of no code but rather process necessary for accessing the API

# Set working directory to Desktop. This can be done using the setwd() function
```{r}
setwd("~/Desktop")
```


# Packages necessary for analysis:
```{r}
library(bitops)
library(RCurl)
library(ROAuth)
library(rjson)
library(streamR)
library(twitteR)
library(base64enc)
```

# Download the certificate for authentication to your desktop
```{r}
download.file(url = "http://curl.haxx.se/ca/cacert.pem", destfile = "cacert.pem")
```

# Create a file to store all the JSON data from the Twitter API
```{r}
outfile <- "tweets_sample.json"
```

# Set up all the configurations to authorize your Twitter application to access the data
```{r}
requestURL <- "https://api.twitter.com/oauth/request_token"
accessURL <- "https://api.twitter.com/oauth/access_token"
authURL <- "https://api.twitter.com/oauth/authorize"
consumerKey <- "hIDz2dvFQ5cBvLOBnwDk9mim3"
consumerSecret <- "tGjbPjzADUFFcvVAe5nA3cwJGJfayEntQqAOB5mAvo5288l39l"
accessToken <- "749969959-6H2Gi16A9i28nVBKAP8cZHJZwHByjd97D5X08Ch"
accessTokenSecret <- "CKxWDP5GTpsKS2HR31vQ5qkZlqRE1koQPMod0kKlIkRrZ"
```

# Allow application to obtain access to the HTTP by authenticating the user via OAuth
# handshake and then save the OAuth certificate locally
```{r}
my_oauth <- OAuthFactory$new(
  consumerKey = consumerKey,
  consumerSecret = consumerSecret,
  requestURL = requestURL,
  accessURL = accessURL,
  authURL = authURL
)

my_oauth$handshake(cainfo = "cacert.pem")
```

# Register the credentials by setting up OAuth credentials for a Twitter session
```{r}
setup_twitter_oauth(consumerKey, consumerSecret, accessToken, accessTokenSecret)
```

